Gaps in stellar streams as a result of globular cluster fly-bys

### 2025-01-13 paper/main.tex
rodrigo's corrects are ingested

    We see clear effects of the presence of their gravitational field in the properties of Palomar~5 stream. A particularly clear example of this effect is shown in Fig.~\ref{fig:stream_on_sky} which we selected for its prominent gaps. Two of these gaps are clearly visible in Galactic coordinates and become even more apparent when marginalizing over latitude to reconstruct the 1D density profile of the stream, as a function of longitude.

@@

    To find out which globular clusters were responsible of creating these gaps, and when close passages took place, we estimated the gravitational acceleration along the orbit of Palomar~5 and represented it in the ($t$, $\tau$) space. $t$ is the simulation time and $\tau$ is indicates how long it will take for Palomar~5 to reach, or how  long ago it passed, a given point on its orbit. The use of $\tau$ is advantageous because the growth of the stream is approximately linear in $\tau$. On the other hand, in physical space streams on eccentric orbits indeed grow in time yet the growth rate is modulated with periodic expansion and contraction depending on the orbital phase ( \citep[see the top panel of Fig.~5.][for an example]{2016MNRAS.457.3817S}). Adopting this time space, and reporting the gravitational acceleration along Palomar~5 orbit in this space, the identification of the globular clusters which produced the perturbation, and the time at which it occurred becomes straight forward. We refer the reader to Appendix~\ref{sec:Perturber_Identification} for an in-depth description of the procedure. In this way, we could identify that the clusters responsible for creating gaps in the simulation of Palomar~5 stream reported in Fig.~\ref{fig:decomposition} are NGC~2808, NGC~7078 and NGC~104, and that their close passages occurred 200~Myr, -1.9 and -2.1~Gyr ago, respectively.

    To further probe that the three clusters above are responsible of producing the gaps observed in this simulation, we conducted additional experiments by including only the perturbation of each of these three clusters at a time, and neglecting the gravitational perturbations of all the other clusters in the Galaxy. The results of these simulations are reported in the middle panels of Fig.~\ref{fig:decomposition} and clearly shows that NGC~2808, NGC~7078 and NGC~104 are the clusters responsible for creating the under-dense regions observed in the leading tails of Palomar~5. It is worth noting that the times at which the passages of these clusters occurred, according to our analysis, are in agreement with the observed width of the corresponding gaps: the encounter with NGC~2808 being very recent (only 200~Myr ago) its induced gap is still very thin, because it takes time for a perturbation to grow into an extended gap, as it is the case for those induced by the passages of  NGC~7078 and NGC~104, which occurred in earlier times. It is also interesting to emphasize that gaps as thin as those generated by the passage of NGC~2808, 200~Myr ago, can be detected by working in the tail coordinate system and by making a comparative analysis (\texttt{full} versus \texttt{reference} simulations): they are so thin that they cannot be directly identified in Galactic coordinates (see Fig.~\ref{fig:stream_on_sky}).\\

@@

    From this analysis we can derive a statistical view of: 1) the number of gaps generated on Pal~5 stream by the whole system of Galactic globular clusters; 2) the clusters which generated these gaps and the time history of these perturbations. From this, we can then quantify: 3) the properties of the perturbers (their masses, sizes and their orbital parameters) as well as 4) the impact geometry of the encounters, which allows us to understand which encounters are more favorable to generate gaps in Pal~5 stream. In the following, we will present results of this analysis, by starting from addressing points 1) and 2).

@@

    By applying the methodology described above to the whole set of simulations, we can reconstruct the history of close passages of Galactic globular clusters to Palomar~5's stream, in the last 5~Gyr, which -- we recall the reader -- is the time interval investigated in our simulations. The results of this analysis are reported in Fig.~\ref{fig:histogram_impact_time} and in Table~\ref{tab:gaps_per_perturber}. NGC~2808 impacts Palomar~5's stream about 200 Myr ago in 44 out of 50 simulations, creating a small gap at a similar position to the one reported in Fig.~\ref{fig:decomposition} in each case. Since this interaction occurred less than one orbital period ago, the uncertainties that lead to orbital solutions remain that similar and thus producing consistent results across all simulations. However, as we continue to turn back time further, the uncertainties in the initial conditions allow the for the various orbital solutions to diverge from one another. Thus, in one configuration a cluster can impact the stream at a given time and yet at the same moment in a different set of initial conditions, it could be on the other side of the galaxy.

@@

    We need a more sophisticated statistic that can be compared to other simulations and turn to the \textit{gap creation rate} developed by \citet{2012ApJ...748...20C}. The gap creation rate is the number of gaps that appear per time and normalized by the length of the stream. For our simulations, this rate is given by: \begin{equation} \label{eq:gap_creation_rate} \mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T}\int_{0}^T l^{-1}(t) \sum_i \delta(t-t_i) dt,\end{equation}where $T$ is the total integration time, $l(t)$ is the length of the stream, and $\sum_i \delta(t-t_i)$ sums over the gap occurrences, with $i$ indexing over the number of gaps in a given simulation. Here, $\delta$ represents the Dirac delta function. This expression can be simplified to:\begin{equation}\mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T} \sum_i \frac{1}{l (t_i)}. \end{equation}

@@

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{impact_geometry_statistics.png}
      \caption{The distribution and relationship between the impact variables from Eq.~\ref{eq:change_in_momentum} for all close fly-bys considered. The encounters that cause gaps are labeled with the same colors as from Fig.~
      \ref{fig:histogram_impact_time} in the scatter plot, and binned in red for the histograms. The close encounters that did not lead to gaps are shown in grey. }
      \label{fig:impact_geometry_statistics}    
    \end{figure*}

    With the perturbers identified, we perform statistics in the pursuit of understanding what conditions are necessary for a globular cluster in order to form a gap on Palomar~5 stream. We turn to impact theory, which in its simplest form is presented in works such as \citet{2008gady.book.....B}. Consider two particles: one stationary and the other moving past it. The distance between the two particles at their closest approach is known as the impact parameter. To simplify the analysis, the \textit{impulse approximation} is employed, which assumes that the velocity of the perturber remains unchanged during the interaction. This assumption simplifies the computation.

@@

    While Fig.~\ref{fig:impact_geometry_statistics} demonstrates that mass, relative velocity, or impact parameter alone cannot predict gap formation, one interesting result emerges: no gaps are created from interactions with impact parameters greater than 300~pc. The stream widths are roughly 200~pc, as seen in Appendix~\ref{sec:gallery_of_gaps}. This finding is even more evident when examining the $b$-$M$ plane. A series of perturbers at roughly $\sim8 \times 10^5 M_\odot$ highlights NGC~2808's flybys, where all encounters with impact parameters under 200~pc result in gaps, while those beyond this distance do not. In other words, even the most massive globulars with masses greater than about $10^6 \rm{M}_\odot$ cannot cause gaps if there impact parameters are greater than about $300$~pc. It is interesting to note that for values of the impact parameter below this threshold, even fast encounters ($\delta v > 300$~km/s) can produce gaps. Perhaps, this is not surprising since the range of possible relative velocities is much less than mass and impact parameter which vary two and three orders, respectively while the relative velocities only vary by about a factor of three.

@@

    \begin{figure*}
      \centering
      \includegraphics[width=0.45\linewidth]{E_Lz_perturbers.png}
      \includegraphics[width=0.45\linewidth]{mass_size_plane.png}
      \includegraphics[width=0.45\linewidth]{impact_stats_phase_space_xy.png}
      \includegraphics[width=0.45\linewidth]{impact_stats_phase_space.png}
      \caption{\textit{Characteristics of gap causing clusters} \textit{Top Left:} Energy-angular momentum space of the globular clusters in the simulations, with 50~$\times$~165 data points representing all sampled initial conditions. Clusters impacting Palomar~5 are shown with colored markers; non-gap-causing clusters are shown in gray. Palomar~5 is marked with a white symbol and black edge. The circular velocity-curve is drawn with a light-pink dashed curve. \textit{Top Right:} Mass-size plane of the globular clusters in the simulations, with uncertainties on the masses indicated as vertical lines. We recall the reader that the globular cluster catalog at this time does not provide uncertainties for clusters' characteristic radii. \textit{Bottom Left:} Palomar~5's orbit in the Galactocentric xy plane. The gray-scale represents all 50 stacked orbits, with $\mathcal{P}$ indicating the probability of  Palomar~5's position, normalized to $\mathcal{P}_\textrm{max}=1$. Markers show the perturbers' positions during each impact. The 50 white dots indicates Palomar~5's sampled initial conditions for the current day position. The colored markers indicate where the perturber was located when it impact the stream and \textit{not} its present day position. \textit{Bottom Right:} same as bottom left but in the meridional plane. In all panels the colors of the markers and histogram bars correspond to specific perturber as specified in Fig.~\ref{fig:histogram_impact_time}.}
      \label{fig:mass_size_plane}
    \end{figure*}

    Once all the key impact parameters estimated, we can use Eq.~\ref{eq:change_in_momentum} and calculate $\Delta p$, the change in momentum (per unit mass) imparted by a cluster flyby on Palomar~5 stream. The result  is shown in Fig.~\ref{fig:deltap}, where the distribution of imparted in change in momentum is shown for all impacts which produce a gap, and compared to those that do not produce one. On average, encounters which lead to gap creations impart a change in momentum on stream particles which is a factor of 10 higher than that of encounters which do not form gaps (but with some overlap in the low-velocity tail). Interestingly, changes in momentum which lead to gap creations extend over a large range in velocities. There is a factor of about 100 between the smallest and largest changes with NGC~2808 (orange color in the histogram) imparting changes in the velocity of stream particles which redistribute over the whole range of $\Delta p$.

    In addition to characterizing the parameters governing cluster encounters with the stream, since we know which clusters have produced gaps on the tail of Palomar~5, we can also verify their orbital and structural properties. This is shown in Fig~\ref{fig:mass_size_plane}, where we first show the mass and size (i.e. half-mass radius) of the clusters that cause gaps on the tails of Palomar~5, dividing them from those that do not. As can be seen, no cluster with mass below $10^5 \rm{M}_\odot$ causes gaps on Palomar~5 stream, and all perturbers, except FSR~1758 have a half-mass radius below 10~pc. Even more interesting is their distribution in the E-Lz plane, which shows that the clusters that cause gaps are on both direct and retrograde orbits (respectively, negative and positive values of $L_z$), but all confined to an energy interval between $-2$ and $-1 \times10^5~\textrm{km}^2\textrm{s}^{-2}$. This is because only clusters within Palomar~5's orbital space can interact with the stream: clusters with higher orbital energies tend to have larger apocenter than that of Palomar~5 and thus spend most of their time away from Palomar~5's orbital volume.

    Finally, it is worth commenting on the location of impacts with the stream, whether they occur when Palomar~5 is close to its pericenter or not. As it can be seen from Fig.~\ref{fig:mass_size_plane} (bottom panel) encounters can occur at all orbital phases of Palomar~5, when it is close to pericenter, but also very far away from it, at the outskirts of its orbital space. However, when taken all together, the location of gap creating impacts shows a strong negative correlation with the galactocentric radius $r$, with the number $N$ of encounters favorable to gap formation  going as $N = -2.5r + 50$ (with a Pearson coefficient of -0.86). While there are more clusters near the Galactic center, clusters naturally spend more time near their apocenters, and the lower relative velocities in these regions should favor gap creation. However, this result suggests that cluster Galactic number density outweighs these factors in determining the number of gaps. It should be noted that these results are for Palomar~5 only. Streams along various orbits needs to be studied in order for this conclusion to be generalized.

@@

    We observe that there are no gaps near within 3~kpc to the cluster within our simulation. There may be a few reasons for the absence of gaps in portion of the tails closer to the cluster center. First, as \citet{2016MNRAS.457.3817S} demonstrated, the dispersion of action-frequencies in the stream plays a role. For instance, the frequency corresponding to the azimuthal action, $J_\phi$, is $\dot\theta_\phi = - \frac{\mathcal{\partial H}}{\partial J_\phi}$ and where $\theta_\phi$ is the angle describing the position of the particle phase space between momentum ($p_\phi$) and position ($\phi$) axes---where $\phi$ is the azimuthal angle between the x-y axes in physical space. Regions of the stream with well-separated frequencies are more susceptible to gap formation, while those with a wide frequency range (near the cluster) tend to erase the history of impacts. Thus, for a gap to form, the imparted change in frequency must exceed the range of frequencies in the impacted region. Thus, the strong flybys that occurred close to the cluster were inconsequential for gap formation.

    Another possible explanation is the different thickness of the simulated tails at different distances from the center of the cluster. At distances of less than about 3~kpc from Palomar~5, the tails are very thin, with a typical thickness being less than 100~pc. In order to cause gaps in these regions, the clusters would have to pass close to the stream, with impact parameters similar to the thickness itself. Note that the thinness of the tails at these distances is probably a direct consequence of the initial parameters we chose for the simulation. We have assumed for Palomar~5, 5~Gyr years ago, the same internal parameters (mass and size) that the cluster has today. With such parameters, after 5~Gyr of evolution, our system has lost most of its mass, and therefore the part of the tails closest to the cluster itself, whose density depends mainly on the most recent mass loss \citep[see, for example, Fig.~A.3 in][]{2012A&A...546L...7M}, is necessarily very thin because the simulated cluster has essentially no more mass to lose. This last point has also a consequence in the gap creation rate, which we derived at the end of Sect.~\ref{sect:geometry}: with only 70\% of the tail  ($\sim$ 14~kpc over 20~kpc, excluding the innermost $\pm 3$~kpc from the cluster center) really suitable for forming gaps, the gap creation rates are about 50$\%$ higher than those estimated in the aforementioned section, where values have been derived taking into account the full tail extent. This gap creation rate would be still not high enough to reproduce the  number of gaps in Palomar~5 tails, as reported by \citet{2012ApJ...760...75C}. In their study, they indeed suggested the presence of  five gaps with a 99~\% detection confidence, leading to a gap creation rate of 0.17~Gyr$^{-1}$~kpc$^{-1}$. If true, this rate would be too high to be explained by Galactic globular clusters only.

@@

  However, this conclusion is not necessarily accurate. Globular clusters lose mass and evaporate over time, leading to an incomplete catalog of perturbers. Moreover, the present-day masses used in our simulations are likely lower than the historical masses of these clusters. For example, \citet{2024ApJ...976...54P} conducted a study simulating the dissolution of a realistic globular cluster population, in the aims of identifying how many stellar streams we should expect in the Milky Way and used a mock catalog of globular clusters with an initial amount with masses above 10~$^4 M_\odot$ totaling at about 10,000 clusters. This implies that more perturbers could have been present in the past, potentially increasing the frequency and number of gaps in Palomar~5's stream.

@@

  As already suggested in previous works which have studied the impact of baryonic structures on Palomar~5 tails \citep{2017NatAs...1..633P, 2019MNRAS.484.2009B}, it is possible that this cluster lies in a region of the phase-space which is not favorable to distinguish gaps created by dark matter sub-halos from gaps created by baryonic structures, as the Galactic bar and giant molecular clouds. With our study, we show that close encounters with globular clusters constitute a further element that confuses a simple interpretation of the observed Palomar~5 stream gaps (if any). Other clusters and streams in the orbital energy range of Palomar~5 may suffer of the same difficulty. Going to lower orbital energies make the situation even worse, because the effect of the Galactic bar, giant molecular clouds and interactions with globular clusters becomes even more efficient. In the innermost regions of the Galaxy, where the density of dark matter sub-halos is expected to be maximal, globular cluster streams are intrinsically more difficult to be found, in addition to the fact that the dynamic times in these regions become so small that the stars lost from globular clusters do not redistribute themselves for the most part into thin structures \citep[see][]{2023A&A...673A..44F}. It is thus probably only at larger distances from the Galactic center than those spanned by Palomar~5 orbit that the impact of dark matter sub-halos may become dominant, but it should be borne in mind that at these distances the number density of sub-halos also decreases. In the future, it will be interesting to apply the type of study conducted here to the whole set of clusters for which streams have been found to quantify the regime in which they are found, whether favorable to the creation of gaps from baryonic structures or not.

@@

  \section{Gap Detection} \label{sec:gap_detection}
    For each of the 50 simulations, we compare the final snapshots between the \texttt{reference} and \texttt{full} simulations using the tail coordinate system, as shown in Fig.~\ref{fig:TailCoordinates}. The \texttt{reference} simulations were only saved at the final timestamps, thus from Eq.~\ref{eq:data_volume_estimate} where $N_ts$=1, leading to a data volume of about one hundred messily megabytes. We inspect these differences by generating 2D density maps of the streams. Additionally, we marginalize over the $y'$-coordinate to produce 1D density profiles along the $x'$-axis. These comparisons are shown in the results section (see Fig.~\ref{fig:profiles}).
    \begin{figure}
      \centering
      \includegraphics[width=\linewidth]{along_orbit_coordinate_system-approved.png}
      \caption{Tail coordinates. \textbf{I want to update this}. }
      \label{fig:TailCoordinates}
    \end{figure}

@@

    To construct the 1D density profiles, we bin the data using the $\sqrt{N}$ rule, where $N$ is the number of data points ($N_p = 10^5$). After binning the 1D profiles, we apply a boxcar smoothing technique. This raw profile density curve is smoothed with the median box-car filter. At each bin, a selected number of adjacent data points from both sides are placed in a list. The bin is then replaced with the median. We choose to use 10 adjacent points per side, which corresponding to a smoothing length of approximately 1~kpc. This procedure reduces high-frequency noise and smooths the profiles. For instance, notice the absence of a high mass peak indicating for the center of mass in the bottom panel of Fig.~\ref{fig:profiles}.

@@

      In this analysis, only one side of the stream is considered. For instance, if the impact candidate was in the leading tail, only the star particles with $\tau > 0$ are used to constrain the stream track. The polynomial coefficients were determined using the through a minimization method using Nelder-Mead algorithm from \texttt{scipy}'s optimization package.

>>>>

    The presence of the other globular clusters has a clear effect on the properties of Palomar~5 stream. A particularly clear example of this effect is shown in Fig.~\ref{fig:stream_on_sky} which we selected for its prominent gaps. Two of these gaps are clearly visible in Galactic coordinates and become even more apparent when marginalizing over latitude to reconstruct the 1D density profile of the stream, as a function of longitude.

@@

    To find out which globular clusters were responsible of creating these gaps, and when close passages took place, we estimated the gravitational acceleration along the orbit of Palomar~5 and represented it in the ($t$, $\tau$) space. $t$ is the simulation time and $\tau$ is indicates how long it will take for Palomar~5 to reach, or how  long ago it passed, a given point on its orbit. The use of $\tau$ is advantageous because the growth of the stream is approximately linear in $\tau$. On the other hand, in physical space streams on eccentric orbits indeed grow in time yet the growth rate is modulated with periodic expansion and contraction depending on the orbital phase \citep[see the top panel of Fig.~5.][for an example]{2016MNRAS.457.3817S}. Adopting this time space, and reporting the gravitational acceleration along Palomar~5 orbit in this space, the identification of the globular clusters which produced the perturbation, and the time at which it occurred becomes straight forward. We refer the reader to Appendix~\ref{sec:Perturber_Identification} for an in-depth description of the procedure. In this way, we could identify that the clusters responsible for creating gaps in the simulation of Palomar~5 stream reported in Fig.~\ref{fig:decomposition} are NGC~2808, NGC~7078 and NGC~104, and that their close passages occurred 200~Myr, -1.9 and -2.1~Gyr ago, respectively.

    To further probe that the three clusters above are responsible of producing the gaps observed in this simulation, we conducted additional experiments by including only the perturbation of each of these three clusters at a time, and neglecting the gravitational perturbations of all the other clusters in the Galaxy. The results of these simulations are reported in the middle panels of Fig.~\ref{fig:decomposition} and clearly show that NGC~2808, NGC~7078 and NGC~104 are the clusters responsible for creating the under-dense regions observed in the leading tails of Palomar~5. It is worth noting that the times at which the passages of these clusters occurred, according to our analysis, are in agreement with the observed width of the corresponding gaps: the encounter with NGC~2808 being very recent (only 200~Myr ago) its induced gap is still very thin, because it takes time for a perturbation to grow into an extended gap, as it is the case for those induced by the passages of  NGC~7078 and NGC~104, which occurred in earlier times. It is also interesting to emphasize that gaps as thin as those generated by the passage of NGC~2808, 200~Myr ago, can be detected by working in the tail coordinate system and by making a comparative analysis (\texttt{full} versus \texttt{reference} simulations): they are so thin that they cannot be directly identified in Galactic coordinates (see Fig.~\ref{fig:stream_on_sky}).\\

@@

    From this analysis we can derive a statistical view of: 1) the number of gaps generated on the Palomar~5 stream by the whole system of Galactic globular clusters; 2) the clusters which generated these gaps and the time history of these perturbations. From this, we can then quantify: 3) the properties of the perturbers (their masses, sizes and their orbital parameters) as well as 4) the impact geometry of the encounters, which allows us to understand which encounters are more favorable to generate gaps in the Palomar~5 stream. In the following, we will present results of this analysis, by starting from addressing points 1) and 2).

@@

    By applying the methodology described above to the whole set of simulations, we can reconstruct the history of close passages of Galactic globular clusters to Palomar~5's stream, in the last 5~Gyr, which -- we remind the reader -- is the time interval investigated in our simulations. The results of this analysis are reported in Fig.~\ref{fig:histogram_impact_time} and in Table~\ref{tab:gaps_per_perturber}. NGC~2808 impacts Palomar~5's stream about 200 Myr ago in 44 out of 50 simulations, creating a small gap at a similar position to the one reported in Fig.~\ref{fig:decomposition} in each case. Since this interaction occurred less than one orbital period ago, despite the uncertainties the orbital solutions remain nontheless similar and thus produce consistent results across all simulations. However, as we continue to turn back time further, the uncertainties in the initial conditions allow the for the various orbital solutions to diverge from one another. Thus, in one configuration a cluster can impact the stream at a given time and yet at the same moment in a different set of initial conditions, it could be on the other side of the galaxy.

@@

    We need a more sophisticated statistic that can be compared to other simulations, so we turn to the \textit{gap creation rate} developed by \citet{2012ApJ...748...20C}. The gap creation rate is the number of gaps that appear per time and normalized by the length of the stream. For our simulations, this rate is given by: \begin{equation} \label{eq:gap_creation_rate} \mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T}\int_{0}^T l^{-1}(t) \sum_i \delta(t-t_i) dt,\end{equation}where $T$ is the total integration time, $l(t)$ is the length of the stream, and $\sum_i \delta(t-t_i)$ sums over the gap occurrences, with $i$ indexing over the number of gaps in a given simulation. Here, $\delta$ represents the Dirac delta function. This expression can be simplified to:\begin{equation}\mathcal{R}_{\textrm{Pal 5}} =  \frac{1}{T} \sum_i \frac{1}{l (t_i)}. \end{equation}

@@

    \begin{figure*}
      \centering
      \includegraphics[width=\linewidth]{impact_geometry_statistics.png}
      \caption{The distribution and relationship between the impact variables from Eq.~\ref{eq:change_in_momentum} for all close fly-bys considered. The encounters that cause gaps are labeled with the same colors as from Fig.~
      \ref{fig:histogram_impact_time} in the scatter plot, and binned in red for the histograms. The close encounters that did not lead to gaps are shown in black. }
      \label{fig:impact_geometry_statistics}    
    \end{figure*}

    With the perturbers identified, we perform statistics in the pursuit of understanding what conditions are necessary for a globular cluster to induce a gap on the Palomar~5 stream. We turn to impact theory, which in its simplest form is presented in works such as \citet{2008gady.book.....B}. Consider two particles: one stationary and the other moving past it. The distance between the two particles at their closest approach is known as the impact parameter. To simplify the analysis, the \textit{impulse approximation} is employed, which assumes that the velocity of the perturber remains unchanged during the interaction. This assumption simplifies the computation.

@@

    While Fig.~\ref{fig:impact_geometry_statistics} demonstrates that mass, relative velocity, or impact parameter alone cannot predict gap formation, one interesting result emerges: no gaps are created from interactions with impact parameters greater than 300~pc. The stream widths are roughly 200~pc, as seen in Appendix~\ref{sec:gallery_of_gaps}. This finding is even more evident when examining the $b$-$M$ plane. A series of perturbers at roughly $\sim8 \times 10^5 M_\odot$ highlights NGC~2808's flybys, where all encounters with impact parameters under 200~pc result in gaps, while those beyond this distance do not. In other words, even the most massive globulars with masses greater than about $10^6 M_\odot$ cannot cause gaps if their impact parameters are greater than about $300$~pc. It is interesting to note that for values of the impact parameter below this threshold, even fast encounters ($\delta v > 300$~km/s) can produce gaps. Perhaps, this is not surprising since the range of possible relative velocities is much less than mass and impact parameter which vary two and three orders, respectively while the relative velocities only vary by about a factor of three.

@@

    \begin{figure*}
      \centering
      \includegraphics[width=0.45\linewidth]{E_Lz_perturbers.png}
      \includegraphics[width=0.45\linewidth]{mass_size_plane.png}
      \includegraphics[width=0.45\linewidth]{impact_stats_phase_space_xy.png}
      \includegraphics[width=0.45\linewidth]{impact_stats_phase_space.png}
      \caption{\textit{Characteristics of gap causing clusters} \textit{Top Left:} Energy-angular momentum space of the globular clusters in the simulations, with 50~$\times$~165 data points representing all sampled initial conditions. Clusters impacting Palomar~5 are shown with colored markers; large for the samplins that induce a gap and small for those that do not. Non-gap-causing clusters are shown in small gray dots. The 50 white dots indicate Palomar~5's sampled initial conditions for the current day position. The circular velocity-curve is drawn with a light-pink dashed curve. \textit{Top Right:} Mass-size plane of the globular clusters in the simulations, with uncertainties on the masses indicated as vertical lines. We recall the reader that the globular cluster catalog at this time does not provide uncertainties for clusters' characteristic radii. \textit{Bottom Left:} Palomar~5's orbit in the Galactocentric xy plane. The gray-scale represents all 50 stacked orbits, with $\mathcal{P}$ indicating the probability of Palomar~5's position, normalized to $\mathcal{P}_\textrm{max}=1$. The colored markers indicate where the perturber was located when it impacted the stream and \textit{not} its present day position. \textit{Bottom Right:} same as bottom left but in the meridional plane. In all panels the colors of the markers and histogram bars correspond to specific perturber as specified in Fig.~\ref{fig:histogram_impact_time}.}
      \label{fig:mass_size_plane}
    \end{figure*}

    Once all the key impact parameters estimated, we can use Eq.~\ref{eq:change_in_momentum} and calculate $\Delta p$, the change in momentum (per unit mass) imparted by a cluster flyby on Palomar~5 stream. The result  is shown in Fig.~\ref{fig:deltap}, where the distribution of imparted change in momentum is shown for all impacts which produce a gap, and compared to those that do not produce one. On average, encounters which lead to gap creations impart a change in momentum on stream particles which is a factor of 10 higher than that of encounters which do not form gaps (but with some overlap in the low-velocity tail). Interestingly, changes in momentum which lead to gap creations extend over a large range in velocities. There is a factor of about 100 between the smallest and largest changes with NGC~2808 (orange color in the histogram) imparting changes in the velocity of stream particles which redistribute over the whole range of $\Delta p$.

    In addition to characterizing the parameters governing cluster encounters with the stream, since we know which clusters have produced gaps on the tail of Palomar~5, we can also verify their orbital and structural properties. This is shown in Fig~\ref{fig:mass_size_plane}, where we first show the mass and size (i.e. half-mass radius) of the clusters that cause gaps on the tails of Palomar~5, dividing them from those that do not. As can be seen, no cluster with mass below $10^5 M_\odot$ causes gaps on the Palomar~5 stream, and all perturbers, except FSR~1758 have a half-mass radius below 10~pc. Even more interesting is their distribution in the E-Lz plane, which shows that the clusters that cause gaps are on both direct and retrograde orbits (respectively, negative and positive values of $L_z$), but all confined to an energy interval between $-2$ and $-1 \times10^5~\textrm{km}^2\textrm{s}^{-2}$. This is because only clusters within Palomar~5's orbital space can interact with the stream: clusters with higher orbital energies tend to have larger apocenter than that of Palomar~5 and thus spend most of their time away from Palomar~5's orbital volume.

    Finally, it is worth commenting on the location of impacts with the stream, whether they occur when Palomar~5 is close to its pericenter or not. As it can be seen from Fig.~\ref{fig:mass_size_plane} (bottom panel) encounters can occur at all orbital phases of Palomar~5, when it is close to pericenter, but also very far away from it, at the outskirts of its orbital space. However, when taken all together, the location of gap creating impacts shows a strong negative correlation with the galactocentric radius $r$, with the number $N$ of encounters favorable to gap formation  going as $N = -2.5r + 50$ (with a Pearson coefficient of -0.86). While there are more clusters near the Galactic center, clusters naturally spend more time near their apocenters, and the lower relative velocities in these regions should favor gap creation. However, this result suggests that cluster Galactic number density outweighs these factors in determining the number of gaps. It should be noted that these results are for Palomar~5 only. Streams along various orbits need to be studied in order for this conclusion to be generalized.

@@

    We observe that there are no gaps near within 3~kpc to the cluster within our simulation. There may be a few reasons for the absence of gaps in the portion of the tails closer to the cluster center. First, as \citet{2016MNRAS.457.3817S} demonstrated, the dispersion of action-frequencies in the stream plays a role. For instance, the frequency corresponding to the azimuthal action, $J_\phi$, is $\dot\theta_\phi = - \frac{\mathcal{\partial H}}{\partial J_\phi}$ and where $\theta_\phi$ is the angle describing the position of the particle phase space between momentum ($p_\phi$) and position ($\phi$) axes---where $\phi$ is the azimuthal angle between the x-y axes in physical space. Regions of the stream with well-separated frequencies are more susceptible to gap formation, while those with a wide frequency range (near the cluster) tend to erase the history of impacts. Thus, for a gap to form, the imparted change in frequency must exceed the range of frequencies in the impacted region. Thus, the strong flybys that occurred close to the cluster were inconsequential for gap formation.

    Another possible explanation is the different thickness of the simulated tails at different distances from the center of the cluster. At distances of less than about 3~kpc from Palomar~5, the tails are very thin, with a typical thickness being less than 100~pc. In order to cause gaps in these regions, the clusters would have to pass close to the stream, with impact parameters similar to the thickness itself. Note that the thinness of the tails at these distances is probably a direct consequence of the initial parameters we chose for the simulation. We have assumed for Palomar~5, 5~Gyr years ago, the same internal parameters (mass and size) that the cluster has today. With such parameters, after 5~Gyr of evolution, our system has lost most of its mass, and therefore the part of the tails closest to the cluster itself, whose density depends mainly on the most recent mass loss \citep[see, for example, Fig.~A.3 in][]{2012A&A...546L...7M}, is necessarily very thin because the simulated cluster has essentially no more mass to lose. This last point has also a consequence in the gap creation rate, which we derived at the end of Sect.~\ref{sect:geometry}: with only 70\% of the tail  ($\sim$ 14~kpc over 20~kpc, excluding the innermost $\pm 3$~kpc from the cluster center) really suitable for forming gaps, the gap creation rates are about 50$\%$ higher than those estimated in the aforementioned section, where values have been derived taking into account the full tail extent. This gap creation rate would be still not high enough to reproduce the  number of gaps in Palomar~5's tails, as reported by \citet{2012ApJ...760...75C}. In their study, they indeed suggested the presence of  five gaps with a 99~\% detection confidence, leading to a gap creation rate of 0.17~Gyr$^{-1}$~kpc$^{-1}$. If true, this rate would be too high to be explained by Galactic globular clusters only.

@@

  However, this conclusion is not necessarily accurate. Globular clusters lose mass and evaporate over time, leading to an incomplete catalog of perturbers. Moreover, the present-day masses used in our simulations are likely lower than the historical masses of these clusters. For example, \citet{2024ApJ...976...54P} conducted a study simulating the dissolution of a realistic globular cluster population, in the aims of identifying how many stellar streams we should expect in the Milky Way and used a mock catalog of globular clusters with an initial amount with masses above 10$^4 M_\odot$ totaling at about 10,000 clusters. This implies that more perturbers could have been present in the past, potentially increasing the frequency and number of gaps in Palomar~5's stream.

@@

  As already suggested in previous works which have studied the impact of baryonic structures on Palomar~5 tails \citep{2017NatAs...1..633P, 2019MNRAS.484.2009B}, it is possible that this cluster lies in a region of the phase-space which is not favorable to distinguish gaps created by dark matter sub-halos from gaps created by baryonic structures, such as the Galactic bar and giant molecular clouds. With our study, we show that close encounters with globular clusters constitute a further element that confuses a simple interpretation of the observed Palomar~5 stream gaps (if any). Other clusters and streams in the orbital energy range of Palomar~5 may suffer of the same difficulty. Going to lower orbital energies make the situation even worse, because the effect of the Galactic bar, giant molecular clouds and interactions with globular clusters becomes even more efficient. In the innermost regions of the Galaxy, where the density of dark matter sub-halos is expected to be maximal, globular cluster streams are intrinsically more difficult to find, in addition to the fact that the dynamic times in these regions become so small that the stars lost from globular clusters do not redistribute themselves for the most part into thin structures \citep[see][]{2023A&A...673A..44F}. It is thus probably only at larger distances from the Galactic center than those spanned by Palomar~5's orbit that the impact of dark matter sub-halos may become dominant, but it should be borne in mind that at these distances the number density of sub-halos also decreases. In the future, it will be interesting to apply the type of study conducted here to the whole set of clusters for which streams have been found to quantify the regime in which they are found, whether favorable to the creation of gaps from baryonic structures or not.

@@

  \section{Gap Detection} \label{sec:gap_detection}
    For each of the 50 simulations, we compare the final snapshots between the \texttt{reference} and \texttt{full} simulations using the tail coordinate system, as shown in Fig.~\ref{fig:TailCoordinates}. The \texttt{reference} simulations were only saved at the final timestamps, thus from Eq.~\ref{eq:data_volume_estimate} where $N_ts$=1, leading to a data volume of about one hundred measly megabytes. We inspect these differences by generating 2D density maps of the streams. Additionally, we marginalize over the $y'$-coordinate to produce 1D density profiles along the $x'$-axis. These comparisons are shown in the results section (see Fig.~\ref{fig:profiles}).
    \begin{figure}
      \centering
      \includegraphics[width=\linewidth]{along_orbit_coordinate_system-approved.png}
      \caption{Tail coordinates. \textbf{I want to update this}. }
      \label{fig:TailCoordinates}
    \end{figure}

@@

    To construct the 1D density profiles, we bin the data using the $\sqrt{N}$ rule, where $N$ is the number of data points ($N_p = 10^5$). After binning the 1D profiles, we apply a boxcar smoothing technique. This raw profile density curve is smoothed with the median box-car filter. At each bin, a selected number of adjacent data points from both sides are placed in a list. The bin is then replaced with the median. We choose to use 10 adjacent points per side, which correspond to a smoothing length of approximately 1~kpc. This procedure reduces high-frequency noise and smooths the profiles. For instance, notice the absence of a high mass peak indicating for the center of mass in the bottom panel of Fig.~\ref{fig:profiles}.

@@

      In this analysis, only one side of the stream is considered. For instance, if the impact candidate was in the leading tail, only the star particles with $\tau > 0$ are used to constrain the stream track. The polynomial coefficients were determined through a minimization method using the Nelder-Mead algorithm from \texttt{scipy}'s optimization package.
